\LARGE \bf Geometric Adaptive Control of Attitude Dynamics on $\SO$\\ with State Inequality Constraints

### 2015-09-24 ACC16.2.tex
first round of proof reading

% discuss issues with parameterizations
All attitude parameterizations fail to represent the nonlinear configuration space both globally and uniquely~\cite{chaturvedi2011a}.
For example, minimal attitude representations, such as Euler angle sequences or modified Rodriguez parameters, suffer from singularities.
These attitude representations are not suitable for large angular slews.
In order to avoid singularities, the designer must carefully switch the chosen Euler angle sequence based on the operating region.
Another option is to artificially limit the operating region of the rigid body.
This ensures the system operates in a region free from singularities but limits the performance capabilities and ability to perform arbitrarily large angular maneuvers.
Quaternions do not have singularities but they double cover the special orthogonal group.
As a result, any physical attitude is represented by a pair of antipodal quaternions on the three-sphere.
An immediate implication of this ambiguity is that closed-loop stability properties derived using quaternions may not hold for the physical rigid body evolving on the special orthogonal group.
During implementation, the design must carefully resolve this non-unique representation in quaternion based attitude control system to avoid undesirable unwinding behavior~\cite{bhat2000}.
For certain initial conditions, where the rigid body starts closed to the desired attitude, the system unnecessarily rotates through a large angle even while the initial error is small.

@@

% discuss how previous papers (mcinnes and lee/meshabi) have short comings
Several approaches have been developed to treat the attitude control problem in the presence of constraints.
A geometric approach is used in~\cite{hablani1999} to determine feasible attitude trajectories prior to implementation.
Typically, an optimal or easily implementable on-board control scheme for attitude maneuvers is considered without state constraints.
The algorithm then seeks an intermediate point such that the control scheme can be applied for each segment without violation of the constraints.
While the geometric method is conceptually simple it is difficult to generalize for an arbitrary amount of constraints.
In addition, this geometric type of approach is only applicable to problems where the selection of intermediate points are computationally feasible

@@

In this work, we develop a geometric adaptive controller on \SO to reorient a rigid body in the presence of state inequality constraints.
We apply a potential function based approach developed directly on the nonlinear manifold \SO.
Geometric control is concerned with the development of control systems for dynamic systems that evolve on nonlinear manifolds.
Artificial potential functions have previously been applied to the attitude control problem~\cite{lee2011b,mcinnes1994}.
These approaches have used attitude parameterizations that give rise to singularities and ambiguities.
By characterizing the intrinsic geometric properties of \SO, we completely avoid issues associated with local parameterizations of the configuration manifold.
In this paper, we present a Lyapunov analysis to guarantee stability in the presence of state constraints.
In addition, we formulate an adaptive update law to enable attitude convergence in the presence of disturbances.
These results are demonstrated via numerical examples and experimental results on a robotic aerial vehicle.
\section{Problem Formulation}\label{sec:prob_form}
\subsection{Attitude Dynamics}\label{sec:att_dyn}
Consider the attitude dynamics of a rigid body. 
Define an inertial reference frame and a body-fixed frame. 
The configuration manifold of the attitude dynamics is the special orthogonal group:
\begin{align*}
	\SO = \{R\in\Re^{3\times 3}\,|\, R^TR=I,\;\mathrm{det}[R]=1\} \, ,
\end{align*}
where a rotation matrix $R\in\SO$ represents the transformation of the representation of a vector from the body-fixed frame to the inertial reference frame. 
The equations of motion are given by
\begin{gather}
	J\dot\Omega + \Omega\times J\Omega = u+W(R,\Omega)\Delta \, ,\label{eqn:Wdot}\\
	\dot R = R\hat\Omega \, ,\label{eqn:Rdot}
\end{gather}
where $J\in\Re^{3\times 3}$ is the inertia matrix, and $\Omega\in\Re^3$ is the angular velocity represented with respect to the body-fixed frame. 
The control moment is denoted by $u\in\Re^3$, and it is expressed with respect to the body-fixed frame. 
We assume that the external disturbance is expressed by $W(R,\Omega)\Delta$, where $W(R,\Omega):\SO\times\Re^3\rightarrow \Re^{3\times p}$ is a known function of the attitude and the angular velocity, and $\Delta\in\Re^p$ is an unknown, but fixed uncertain parameter.
In addition, we assume that a bound on \( W(R, \Omega) \) is known and given by
\begin{equation}
	\norm{W} \leq B_W \, . \label{eqn:W_bound}
\end{equation}

In~\refeqn{Rdot}, the \textit{hat} map $\wedge :\Re^{3}\rightarrow\so$ represents the transformation of a vector in $\Re^3$ to a $3\times 3$ skew-symmetric matrix such that $\hat x y = x\times y$ for any $x,y\in\Re^3$~\cite{BulLew05}. More explicitly, 
\begin{align*}
\hat x = \begin{bmatrix} 0 & -x_3 & x_2 \\ x_3 & 0 & -x_1 \\ -x_2 & x_1 & 0\end{bmatrix},
\end{align*}
for $x=[x_1,x_2,x_3]^T\in\Re^3$. 
The inverse of the hat map is denoted by the \textit{vee} map $\vee:\so\rightarrow\Re^3$. Several properties of the hat map are summarized as
\begin{gather}
%    \hat x y = x\times y = - y\times x = - \hat y x,\\
    x\cdot \hat y z = y\cdot \hat z x,\quad \hat x\hat y z = (x\cdot z) y - (x\cdot y ) z\label{eqn:STP},\\
%    \hat x\hat y z = (x\cdot z) y - (x\cdot y ) z,\label{eqn:VTP}\\
%    C^TC=-C^T\hat e_3^2 C = I_2
%    \hat x^T \hat x = (x^T x) I - x x^T,\\
%    \hat x \hat y \hat x=-(y^Tx)\hat x,\\
%    -\frac{1}{2}\tr{\hat x \hat y} = x^T y,\\
    \widehat{x\times y} = \hat x \hat y -\hat y \hat x = yx^T-xy^T,\label{eqn:hatxy}\\
    %\tr{\hat x A}=
    \tr{A\hat x }=\frac{1}{2}\tr{\hat x (A-A^T)}=-x^T (A-A^T)^\vee,\label{eqn:hat1}\\
%    \widehat{Ax} = \hat x \parenth{\frac{1}{2}\tr{A}I-A} + \parenth{\frac{1}{2}\tr{A}I-A}^T \hat x,\\
    \hat x  A+A^T\hat x=(\braces{\tr{A}I_{3\times 3}-A}x)^{\wedge},\label{eqn:xAAx}\\
R\hat x R^T = (Rx)^\wedge,\quad 
R(x\times y) = Rx\times Ry\label{eqn:RxR}
%,\label{eqn:Rxy}
\end{gather}
for any $x,y,z\in\Re^3$, $A\in\Re^{3\times 3}$ and $R\in\SO$. 
Throughout this paper, the dot product of two vectors is denoted by $x\cdot y = x^T y$ for any $x,y\in\Re^n$ and the maximum eigenvalue and the minimum eigenvalue of $J$ are denoted by $\lambda_M$ and $\lambda_m$, respectively. 
The 2-norm of a matrix \( A \) is denoted by \( \norm{A} \), and its Frobenius norm is denoted by \( \norm{A}_F = \tr{A^T A} \).
This results in \( \norm{A} \leq \norm{A}_F \leq \sqrt{r}\norm{A} \), where \( r \) is the rank of A.

@@

The two-sphere is the manifold of unit-vectors in \( \R^3 \) such that \( \Sph^2 = \{ q \in \R^3 \,  \vert \, \norm{q} = 1 \}\).
We define \( r \in \Sph^2 \) to be a unit vector from the mass center of the rigid body and represented with respect to the body-fixed frame.
For example, \( r \) may represent the pointing direction of an on board optical sensor.
We define \( v \in \Sph^2 \) to be a unit vector from the mass center of the rigid body towards an undesired pointing direction and represented in the inertial reference frame.
For example, \( v \) may represent the inertial direction of a bright celestial object or the incoming direction of particles or other debris
It is further assumed that optical sensor has a strict non-exposure constraint with respect to the celestial object.
We formulate this hard constraint as
\begin{align}
	r^T R^T v \leq \cos \theta \, , \label{eqn:constraint}
\end{align}
where \( \ang{0} \leq \theta \leq \ang{90}  \) is the required minimum angular separation between \( r \) and \( R^T v \). 
The objective is to a determine a control input \( u \) that stabilizes the system from an initial attitude \( R_0 \) to a desired attitude \( R_d \) while ensuring that~\cref{eqn:constraint} is always satisfied.

\section{Attitude Control on $\SO$ with Inequality Constraints}
The first step in designing a control system on a nonlinear manifold \( \Q \) is the selection of a proper configuration error function.
This configuration error function, \( \Psi : \Q \times \Q \to \R \), is a smooth and proper positive definite function that measures the error between the current configuration and a desired configuration.
Once an appropriate configuration error function is chosen, one can then define a configuration error vector and a velocity error vector in the tangent space \( \mathsf{T}_q \Q \) through use of the derivatives of \( \Psi \)~\cite{bullo2004}.
With the configuration error function and vectors the remaining procedure is similar to nonlinear control design on Euclidean vector spaces.
One chooses control inputs as functions of the state through a Lyapunov analysis on \Q.

A new configuration error function is proposed for attitude control in the presence of state inequality constraints.
We extend the form used in~~\cite{bullo2004,lee2011a} for attitude control on \SO with the addition of a logarithmic barrier function.
We use the configuration error function to develop a nonlinear geometric attitude controller for the stabilization of a general rigid body. 
A smooth control system is first developed for the no disturbance case and then extended to include an adaptive update law for stabilization in the presence of unknown disturbances.

\begin{prop}[Attitude Error Function] \label{prop:config_error}
We assume a desired attitude command \( ( R_d, \Omega_d = 0 ) \), the current attitude and angular velocity \( ( R, \Omega) \), and a state constraint~\refeq{eqn:constraint} are given.
We define an attitude error function \( \Psi : \SO \times \SO \to \R \), and attitude error vector \( e_R : \SO \times \SO \to \R \), and an angular velocity error vector \( e_\Omega : \R^3 \to \R^3 \) as follows:
\begin{gather}
	\Psi(R) = A(R) B(R) \, , \label{eqn:psi} \\
	e_R = e_{R_A} B + A e_{R_B} \, , \label{eqn:eR} \\
	e_\Omega = \Omega \, , \label{eqn:eW}
\end{gather}
with
\begin{gather}
	A(R) = \frac{1}{2} \tr{G \left( I - R_d^T R\right)} \, , \label{eqn:A} \\
	B(R) = 1 - \frac{1}{\alpha} \ln \left( \frac{\cos \theta -  r^T R^T v}{1 + \cos \theta}\right) \, . \label{eqn:B} \\
	e_{R_A} = \frac{1}{2} \parenth{G R_d^T R - R^T R_d G} ^ \vee \, , \label{eqn:eRA} \\
	e_{R_B} = \frac{ \left( R^T v\right)^\vee r}{\alpha \left(r^T R^T v - \cos \theta \right)} \, . \label{eqn:eRB} 
\end{gather}	
where the matrix \( G \in \R^{3 \times 3} \) is defined as a diagonal matrix matrix for distinct, positive constants \( g_1, g_2, g_3 \in \R \).
Then, the following properties hold
\begin{enumerate}[(i)]
	\item \label{item:prop_psi_psd} \(\Psi\) is locally positive definite about \( R = R_d\)
	\item \label{item:prop_er} The variation of \( \Psi \) with respect to a variation of \( \delta R = R \hat{\eta} \) for \( \eta \in \R^3 \) is given by
	\begin{align}
		\dirDiff{\Psi}{R} &= \dirDiff{A}{R} \,  B(R) + A \,  \dirDiff{B}{R} \, . \\
	\end{align}
	\item \label{item:prop_era}The variation of \( A(R) \) with respect to a variation of \( \delta R = R \hat{\eta} \) for \( \eta \in \R^3 \) is given by
	\begin{align}
		\dirDiff{A}{R} &= \eta \cdot e_{R_A} \, .
	\end{align}
	\item \label{item:prop_erb} The variation of \( B(R) \) with respect to a variation of \( \delta R = R \hat{\eta} \) for \( \eta \in \R^3 \) is given by
	\begin{align}
		\dirDiff{B_i}{R} &= \eta \cdot e_{R_{B_i}} \, .
	\end{align}
	\item \label{item:prop_crit}The critical points of \( \Psi \),where \( e_R = 0\), are \( \braces{R_d} \cup \braces{R_d \exp(\pi \hat{s}} \) for \( s \in \braces{e_1, e_2, e_3} \) and \( R^T v = \pm r \).
	\item \label{item:prop_era_upbound}An upper bound of \( \norm{e_{R_A}} \) is given as:
	\begin{align}
		\norm{e_{R_A}}^2 \leq \frac{\Psi}{b_1} \, , \label{eqn:psi_lower_bound}
	\end{align}
	where the constant \( b_1 \) is given by \( b_1 = \frac{h_1}{h_2 + h_3} \) for 
	\begin{align*}
		h_1 &= \min{g_1 + g_2, g_2 + g_3 , g_3 + g_1} \, ,\\
		h_2 &= \min{\parenth{g_1 -g_2}^2,\parenth{g_2 -g_3}^2 , \parenth{g_3 -g_1}^2} \, ,\\
		h_3 &= \min{\parenth{g_1 + g_2}^2, \parenth{g_2 + g_3}^2 , \parenth{g_3 + g_1}^2} \, .				
	\end{align*}
\end{enumerate}
\end{prop}
\begin{proof}[Proof of~\Cref{prop:config_error}]\label{proof:config_error}
The proof of~\cref{item:prop_era_upbound} is available in~\cite{lee2011a}.
To prove~\cref{item:prop_psi_psd} we note that~\cref{eqn:A} is a locally positive definite function about \( R = R_d \)~\cite{bullo2004}.
The constraint angle is assumed \( \ang{0} \leq \theta \leq \ang{90} \) such that \( 0 \leq \cos \theta \).
The term \( r^T R^T v \) represents the cosine of the angle between the body fixed vector \( r \) and the inertial vector \( v \). 
It follows that
\begin{align*}
	0 \leq  \frac{\cos \theta -  r^T R^T v}{1 + \cos \theta} \leq 1 \, ,
\end{align*}
for all \( R \in SO \). 
As a result, the negative logarithm is always positive and
\begin{align*}
	1 < B \, .
\end{align*}
The product~\cref{eqn:psi} is composed of two positive terms and is therefore also positive definite, and has a single critical point at \( R = R_d \).

@@

A straightforward application of the chain and product rules of differentiation allows us to show~\cref{item:prop_erb} as
\begin{align*}
	\dirDiff{B}{R} &= -\frac{1}{\alpha} \left[\left(\frac{1+\cos \theta}{\cos \theta - r^T R^T v} \right) \left(\eta \cdot \frac{\left( R^T v\right)^\vee r}{1+\cos \theta_i} \right) \right]  \\
	&= \eta \cdot \frac{ - \left( R^T v\right)^\vee r}{\alpha \left(\cos \theta - r^T R^T v \right)} \, ,
\end{align*}
where the scalar triple product~\cref{eqn:STP} was used.

To show~\cref{item:prop_crit}, we determine the critical points of~\cref{eqn:eRA}.
The attitude error between \( R \) and \( R_d \) is defined as \( Q \in SO \) as
\begin{align*}
	Q = R_d^T R = 
	\begin{bmatrix}
		q_{11} & q_{12} & q_{13} \\
		q_{21} & q_{22} & q_{23} \\
		q_{31} & q_{32} & q_{33}
	\end{bmatrix} \, .
\end{align*}
Equating \( e_{R_{A}} = 0 \) yields the following equivalent relations
\begin{subequations}
\begin{align}
	g_3 q_{32} &= g_2 q_{23} \, , \\
	g_1 q_{13} &= g_3 q_{31} \, , \\
	g_2 q_{21} &= g_1 q_{12} \, .
\end{align}
\end{subequations}
These relations allow one to redefine the attitude error \( Q \) in terms of the independent upper diagonal elements.
Noting that \( Q \in \SO \) is an orthonormal matrix, with unit length row and columns gives
\begin{subequations}
\begin{align}
	\parenth{1 - \frac{g_1^2}{g_2^2}} q_{12}^2 + \parenth{1 - \frac{g_1^2}{g_3^2}}q_{13}^2 &= 0 \, , \\
	\parenth{1 - \frac{g_1^2}{g_2^2}} q_{12}^2 - \parenth{1 - \frac{g_2^2}{g_3^2}} q_{23}^2 &= 0 \, , \\
	\parenth{1-\frac{g_1^2}{g_3^2}} q_{13}^2 + \parenth{1 - \frac{g_2^2}{g_3^2}} q_{23}^2 &= 0 \, .
\end{align}
\end{subequations}
As a result, all the off diagonal terms of \( Q = R_d^T R \), namely \( q_{12}= q_{13} = q_{23} = 0 \).
Since \( Q \in \SO \), the diagonal elements \( q_{11}, q_{22}, \text{ and } q_{33} \) are either \( +1 \) or \(-1\).
There are four critical points of \( e_{R_A} \), the desired attitude \( R_d \) as well as rotations about each body fixed axis by \ang{180}.
The repulsive error vector \( e_{R_{B}} \) is zero only when the numerator \( \parenth{R^T v}^\wedge r = 0 \). 
This condition only occurs if the desired attitude results in the body fixed vector \( r \) becoming aligned with \(R^T v \) while simultaneously satisfying \(\braces{R_d} \cup \braces{R_d \exp(\pi \hat{s}} \) for \( s \in \braces{e_1, e_2, e_3} \) .
The desired equilibrium is \( e_R = 0 \) and \( A = 0\).
\end{proof}

>>>>

% discuss issues with parameterizations
All attitude parameterizations fail to represent the nonlinear configuration space both globally and uniquely~\cite{chaturvedi2011a}.
For example, minimal attitude representations, such as Euler angle sequences or modified Rodriguez parameters, suffer from singularities.
These attitude representations are not suitable for large angular slews.
In order to avoid singularities, the designer must carefully switch the chosen Euler angle sequence based on the operating region.
Another option is to artificially limit the operating region of the rigid body.
This ensures the system operates in a region free from singularities but limits the performance capabilities and ability to perform arbitrarily large angular maneuvers.
Quaternions do not have singularities but they double cover the special orthogonal group.
As a result, any physical attitude is represented by a pair of antipodal quaternions on the three-sphere.
An immediate implication of this ambiguity is that closed-loop stability properties derived using quaternions may not hold for the physical rigid body evolving on the special orthogonal group.
During implementation, the designer must carefully resolve this non-unique representation in quaternion based attitude control systems to avoid undesirable unwinding behavior~\cite{bhat2000}.
Unwinding is where the rigid body starts close to the desired attitude, yet the system unnecessarily rotates through a large angle even while the initial error is small.

@@

% discuss how previous papers (mcinnes and lee/meshabi) have short comings
Several approaches have been developed to treat the attitude control problem in the presence of constraints.
A geometric approach is used in~\cite{hablani1999} to determine feasible attitude trajectories prior to implementation.
Typically, an optimal or easily implementable on-board control scheme for attitude maneuvers is considered without state constraints.
The algorithm then seeks an intermediate point such that the control scheme can be applied for each segment without violation of the constraints.
While the geometric method is conceptually simple it is difficult to generalize for an arbitrary number of constraints.
In addition, this geometric type of approach is only applicable to problems where the selection of intermediate points are computationally feasible

@@

In this work, we develop a geometric adaptive controller on \(\SO\) to reorient a rigid body in the presence of state inequality constraints.
We apply a potential function based approach developed directly on the nonlinear manifold \SO.
Geometric control is concerned with the development of control systems for dynamic systems that evolve on nonlinear manifolds.
Artificial potential functions have previously been applied to the attitude control problem~\cite{lee2011b,mcinnes1994}.
These approaches have used attitude parameterizations that give rise to singularities and ambiguities.
By characterizing the intrinsic geometric properties of \SO, we completely avoid issues associated with local parameterizations of the configuration manifold.
In this paper, we present a Lyapunov analysis to guarantee stability in the presence of state constraints.
In addition, we formulate an adaptive update law to enable attitude convergence in the presence of disturbances.
These results are demonstrated via numerical examples and experimental results on a hexrotor aerial vehicle.
\section{Problem Formulation}\label{sec:prob_form}
\subsection{Attitude Dynamics}\label{sec:att_dyn}
Consider the attitude dynamics of a rigid body. 
We define an inertial reference frame and a body-fixed frame fixed to the center of mass. 
The configuration manifold of the attitude dynamics is the special orthogonal group:
\begin{align*}
	\SO = \{R\in\Re^{3\times 3}\,|\, R^TR=I,\;\mathrm{det}[R]=1\} \, ,
\end{align*}
where a rotation matrix $R\in\SO$ represents the transformation of the representation of a vector from the body-fixed frame to the inertial reference frame. 
The equations of motion are given by
\begin{gather}
	J\dot\Omega + \Omega\times J\Omega = u+W(R,\Omega)\Delta \, ,\label{eqn:Wdot}\\
	\dot R = R\hat\Omega \, ,\label{eqn:Rdot}
\end{gather}
where $J\in\R^{3\times 3}$ is the inertia matrix, and $\Omega\in\R^{3\times1}$ is the angular velocity represented with respect to the body-fixed frame. 
The control moment is denoted by $u\in\R^{3\times1}$, and it is expressed with respect to the body-fixed frame. 
We assume that the external disturbance is expressed by $W(R,\Omega)\Delta$, where $W(R,\Omega):\SO\times\R^{3\times1}\rightarrow \R^{3\times p}$ is a known function of the attitude and the angular velocity.
The disturbance is represented by $\Delta\in\R^{p\times1}$ is an unknown, but fixed uncertain parameter.
In addition, we assume that a bound on \( W(R, \Omega) \) is known and given by
\begin{equation}
	\norm{W} \leq B_W \, . \label{eqn:W_bound}
\end{equation}

In~\refeqn{Rdot}, the \textit{hat} map $\wedge :\R^{3\times1}\rightarrow\so$ represents the transformation of a vector in $\R^{3\times1}$ to a $3\times 3$ skew-symmetric matrix such that $\hat x y = x\times y$ for any $x,y\in\R^{3\times1}$~\cite{bullo2004}. 
More explicitly, 
\begin{align*}
\hat x = \begin{bmatrix} 0 & -x_3 & x_2 \\ x_3 & 0 & -x_1 \\ -x_2 & x_1 & 0\end{bmatrix},
\end{align*}
for $x=[x_1,x_2,x_3]^T\in\R^{3\times1}$. 
The inverse of the hat map is denoted by the \textit{vee} map $\vee:\so\rightarrow\R^{3\times1}$. 
Several properties of the hat map are summarized as
\begin{gather}
%    \hat x y = x\times y = - y\times x = - \hat y x,\\
    x\cdot \hat y z = y\cdot \hat z x,\quad \hat x\hat y z = (x\cdot z) y - (x\cdot y ) z\label{eqn:STP},\\
%    \hat x\hat y z = (x\cdot z) y - (x\cdot y ) z,\label{eqn:VTP}\\
%    C^TC=-C^T\hat e_3^2 C = I_2
%    \hat x^T \hat x = (x^T x) I - x x^T,\\
%    \hat x \hat y \hat x=-(y^Tx)\hat x,\\
%    -\frac{1}{2}\tr{\hat x \hat y} = x^T y,\\
    \widehat{x\times y} = \hat x \hat y -\hat y \hat x = yx^T-xy^T,\label{eqn:hatxy}\\
    %\tr{\hat x A}=
    \tr{A\hat x }=\frac{1}{2}\tr{\hat x (A-A^T)}=-x^T (A-A^T)^\vee,\label{eqn:hat1}\\
%    \widehat{Ax} = \hat x \parenth{\frac{1}{2}\tr{A}I-A} + \parenth{\frac{1}{2}\tr{A}I-A}^T \hat x,\\
    \hat x  A+A^T\hat x=(\braces{\tr{A}I_{3\times 3}-A}x)^{\wedge},\label{eqn:xAAx}\\
R\hat x R^T = (Rx)^\wedge,\quad 
R(x\times y) = Rx\times Ry\label{eqn:RxR}
%,\label{eqn:Rxy}
\end{gather}
for any $x,y,z\in\R^{3\times1}$, $A\in\R^{3\times 3}$ and $R\in\SO$. 
Throughout this paper, the dot product of two vectors is denoted by $x\cdot y = x^T y$ for any $x,y\in\R^n$ and the maximum eigenvalue and the minimum eigenvalue of $J$ are denoted by $\lambda_M$ and $\lambda_m$, respectively. 
The 2-norm of a matrix \( A \) is denoted by \( \norm{A} \), and its Frobenius norm is denoted by \( \norm{A}_F = \tr{A^T A} \).
This results in \( \norm{A} \leq \norm{A}_F \leq \sqrt{r}\norm{A} \), where \( r \) is the rank of A.

@@

The two-sphere is the manifold of unit-vectors in \( \R^3 \) such that \( \Sph^2 = \{ q \in \R^3 \,  \vert \, \norm{q} = 1 \}\).
We define \( r \in \Sph^2 \) to be a unit vector from the mass center of the rigid body and represented with respect to the body-fixed frame.
For example, \( r \) may represent the pointing direction of an on-board optical sensor.
We define \( v \in \Sph^2 \) to be a unit vector from the mass center of the rigid body towards an undesired pointing direction and represented in the inertial reference frame.
For example, \( v \) may represent the inertial direction of a bright celestial object or the incoming direction of particles or other debris.
It is further assumed that optical sensor has a strict non-exposure constraint with respect to the celestial object.
We formulate this hard constraint as
\begin{align}
	r^T R^T v \leq \cos \theta \, , \label{eqn:constraint}
\end{align}
where \( \ang{0} \leq \theta \leq \ang{90}  \) is the required minimum angular separation between \( r \) and \( R^T v \). 
The objective is to a determine a control input \( u \) that stabilizes the system from an initial attitude \( R_0 \) to a desired attitude \( R_d \) while ensuring that~\cref{eqn:constraint} is always satisfied.

\section{Attitude Control on $\SO$ with Inequality Constraints}
The first step in designing a control system on a nonlinear manifold \( \Q \) is the selection of a proper configuration error function.
This configuration error function, \( \Psi : \Q \times \Q \to \R \), is a smooth and proper positive definite function that measures the error between the current configuration and a desired configuration.
Once an appropriate configuration error function is chosen, one can then define a configuration error vector and a velocity error vector in the tangent space \( \mathsf{T}_q \Q \) through use of the derivatives of \( \Psi \)~\cite{bullo2004}.
With the configuration error function and vectors the remaining procedure is analogous to nonlinear control design on Euclidean vector spaces.
One chooses control inputs as functions of the state through a Lyapunov analysis on \Q.

A new configuration error function is proposed for attitude control in the presence of state inequality constraints.
We extend the form used in~\cite{bullo2004,lee2011a} for attitude control on \SO with the addition of a logarithmic barrier function.
We use the configuration error function to develop a nonlinear geometric attitude controller for the stabilization of a general rigid body. 
A smooth control system is first developed for the no disturbance case and then extended to include an adaptive update law for stabilization in the presence of unknown disturbances.

\begin{prop}[Attitude Error Function] \label{prop:config_error}
We assume a desired attitude command \( ( R_d, \Omega_d = 0 ) \), the current attitude and angular velocity \( ( R, \Omega) \), and a state constraint~\cref{eqn:constraint} are given.
We define an attitude error function \( \Psi : \SO \times \SO \to \R \), and attitude error vector \( e_R : \SO \times \SO \to \R \), and an angular velocity error vector \( e_\Omega : \R^3 \to \R^3 \) as follows:
\begin{gather}
	\Psi(R) = A(R) B(R) \, , \label{eqn:psi} \\
	e_R = e_{R_A} B + A e_{R_B} \, , \label{eqn:eR} \\
	e_\Omega = \Omega \, , \label{eqn:eW}
\end{gather}
with
\begin{gather}
	A(R) = \frac{1}{2} \tr{G \left( I - R_d^T R\right)} \, , \label{eqn:A} \\
	B(R) = 1 - \frac{1}{\alpha} \ln \left( \frac{\cos \theta -  r^T R^T v}{1 + \cos \theta}\right) \, . \label{eqn:B} \\
	e_{R_A} = \frac{1}{2} \parenth{G R_d^T R - R^T R_d G} ^ \vee \, , \label{eqn:eRA} \\
	e_{R_B} = \frac{ \left( R^T v\right)^\vee r}{\alpha \left(r^T R^T v - \cos \theta \right)} \, . \label{eqn:eRB} 
\end{gather}	
where the matrix \( G \in \R^{3 \times 3} \) is defined as a diagonal matrix matrix for distinct, positive constants \( g_1, g_2, g_3 \in \R \).
Then, the following properties hold
\begin{enumerate}[(i)]
	\item \label{item:prop_psi_psd} \(\Psi\) is locally positive definite about \( R = R_d\)
	\item \label{item:prop_er} The variation of \( \Psi \) with respect to a variation of \( \delta R = R \hat{\eta} \) for \( \eta \in \R^3 \) is given by
	\begin{align}
		\dirDiff{\Psi}{R} &= \dirDiff{A}{R} \,  B(R) + A \,  \dirDiff{B}{R} \, . \\
	\end{align}
	\item \label{item:prop_era}The variation of \( A(R) \) with respect to a variation of \( \delta R = R \hat{\eta} \) for \( \eta \in \R^3 \) is given by
	\begin{align}
		\dirDiff{A}{R} &= \eta \cdot e_{R_A} \, .
	\end{align}
	\item \label{item:prop_erb} The variation of \( B(R) \) with respect to a variation of \( \delta R = R \hat{\eta} \) for \( \eta \in \R^3 \) is given by
	\begin{align}
		\dirDiff{B}{R} &= \eta \cdot e_{R_{B}} \, .
	\end{align}
	\item \label{item:prop_crit}The critical points of \( \Psi \),where \( e_R = 0\), are \( \braces{R_d} \cup \braces{R_d \exp(\pi \hat{s}} \) for \( s \in \braces{e_1, e_2, e_3} \) and \( R^T v = \pm r \).
	\item \label{item:prop_era_upbound}An upper bound of \( \norm{e_{R_A}} \) is given as:
	\begin{align}
		\norm{e_{R_A}}^2 \leq \frac{\Psi}{b_1} \, , \label{eqn:psi_lower_bound}
	\end{align}
	where the constant \( b_1 \) is given by \( b_1 = \frac{h_1}{h_2 + h_3} \) for 
	\begin{align*}
		h_1 &= \min{g_1 + g_2, g_2 + g_3 , g_3 + g_1} \, ,\\
		h_2 &= \min{\parenth{g_1 -g_2}^2,\parenth{g_2 -g_3}^2 , \parenth{g_3 -g_1}^2} \, ,\\
		h_3 &= \min{\parenth{g_1 + g_2}^2, \parenth{g_2 + g_3}^2 , \parenth{g_3 + g_1}^2} \, .				
	\end{align*}
\end{enumerate}
\end{prop}
\begin{proof}[Proof of~\Cref{prop:config_error}]\label{proof:config_error}
The proof of~\cref{item:prop_era_upbound} is available in~\cite{lee2011a}.
To prove~\cref{item:prop_psi_psd} we note that~\cref{eqn:A} is a locally positive definite function about \( R = R_d \)~\cite{bullo2004}.
The constraint angle is assumed \( \ang{0} \leq \theta \leq \ang{90} \) such that \( 0 \leq \cos \theta \).
The term \( r^T R^T v \) represents the cosine of the angle between the body fixed vector \( r \) and the inertial vector \( v \). 
It follows that
\begin{align*}
	0 \leq  \frac{\cos \theta -  r^T R^T v}{1 + \cos \theta} \leq 1 \, ,
\end{align*}
for all \( R \in \SO \). 
As a result, the negative logarithm is always positive and
\begin{align*}
	1 < B \, .
\end{align*}
The product~\cref{eqn:psi} is composed of two positive terms and is therefore also positive definite, and has a single critical point at \( R = R_d \).

@@

A straightforward application of the chain and product rules of differentiation allows us to show~\cref{item:prop_erb} as
\begin{align*}
	\dirDiff{B}{R} &= -\frac{1}{\alpha} \left[\left(\frac{1+\cos \theta}{\cos \theta - r^T R^T v} \right) \left(\eta \cdot \frac{\left( R^T v\right)^\vee r}{1+\cos \theta} \right) \right]  \\
	&= \eta \cdot \frac{ - \left( R^T v\right)^\vee r}{\alpha \left(\cos \theta - r^T R^T v \right)} \, ,
\end{align*}
where the scalar triple product~\cref{eqn:STP} was used.

To show~\cref{item:prop_crit}, we determine the critical points of~\cref{eqn:eRA}.
The attitude error between \( R \) and \( R_d \) is defined by \( Q \in \SO \) as
\begin{align*}
	Q = R_d^T R = 
	\begin{bmatrix}
		q_{11} & q_{12} & q_{13} \\
		q_{21} & q_{22} & q_{23} \\
		q_{31} & q_{32} & q_{33}
	\end{bmatrix} \, .
\end{align*}
Equating \( e_{R_{A}} = 0 \) yields the following equivalent relations
\begin{subequations}
\begin{align}
	g_3 q_{32} &= g_2 q_{23} \, , \\
	g_1 q_{13} &= g_3 q_{31} \, , \\
	g_2 q_{21} &= g_1 q_{12} \, .
\end{align}
\end{subequations}
These relations allow one to redefine the attitude error \( Q \) in terms of the independent upper diagonal elements.
Noting that \( Q \in \SO \) is an orthonormal matrix, with unit length row and columns gives
\begin{subequations}
\begin{align}
	\parenth{1 - \frac{g_1^2}{g_2^2}} q_{12}^2 + \parenth{1 - \frac{g_1^2}{g_3^2}}q_{13}^2 &= 0 \, , \\
	\parenth{1 - \frac{g_1^2}{g_2^2}} q_{12}^2 - \parenth{1 - \frac{g_2^2}{g_3^2}} q_{23}^2 &= 0 \, , \\
	\parenth{1-\frac{g_1^2}{g_3^2}} q_{13}^2 + \parenth{1 - \frac{g_2^2}{g_3^2}} q_{23}^2 &= 0 \, .
\end{align}
\end{subequations}
As a result, all the off diagonal terms of \( Q = R_d^T R \) are zero, namely \( q_{12}= q_{13} = q_{23} = 0 \).
Since \( Q \in \SO \), the diagonal elements \( q_{11}, q_{22}, \text{ and } q_{33} \) are either \( +1 \) or \(-1\).
There are four critical points of \( e_{R_A} \), the desired attitude \( R_d \) as well as rotations about each body fixed axis by \ang{180}.
The repulsive error vector \( e_{R_{B}} \) is zero only when the numerator \( \parenth{R^T v}^\wedge r = 0 \). 
This condition only occurs if the desired attitude results in the body fixed vector \( r \) becoming aligned with \(R^T v \) while simultaneously satisfying \(\braces{R_d} \cup \braces{R_d \exp(\pi \hat{s}} \) for \( s \in \braces{e_1, e_2, e_3} \) .
The desired equilibrium is \( e_R = 0 \) and \( A = 0\).
\end{proof}
